Supermartingales and Termination Analysis of Statistical PCF

### 2021-01-05 LICS21/antitone.tex
Sec. VI. Ex. VI.2

\changed[akr]{Note that antitone ranking functions are actually a generalisation of ranking functions, even though their name may suggest that they are a type of ranking functions.}

@@

%\lo{The preceding definition does not place any constraint on the stopping time $T$ (other than that it is w.r.t.~the filtration $(\calF_n)_{n \geq 0}$). However for the definition to make sense, $T$ must be the stopping time $T(s) := \min \set{n \mid Y_n(s) = 0}$.}
\akr{The definition you gave for an antitone ranking function is incorrect, because the antitone ranking functions are a proper subset of ranking functions. Although we may have been referring to antitone ranking functions as a type of ranking function, they're a generalisation, so they don't actually meet the definition of a ranking function in general.}

>>>>

\changed[lo]{Note that antitone ranking functions are actually a generalisation of ranking functions, even though the way we reference them may suggest that they are a type of ranking functions.}

@@

%\lo{The preceding definition does not place any constraint on the stopping time $T$ (other than that it is w.r.t.~the filtration $(\calF_n)_{n \geq 0}$). However for the definition to make sense, $T$ must be the stopping time $T(s) := \min \set{n \mid Y_n(s) = 0}$.}
%\akr{The definition you gave for an antitone ranking function is incorrect, because the antitone ranking functions are a proper subset of ranking functions. Although we may have been referring to antitone ranking functions as a type of ranking function, they're a generalisation, so they don't actually meet the definition of a ranking function in general.}

### 2021-01-05 LICS21/confluent.tex
Sec. VI. Ex. VI.2

A \emph{position} is a finite sequence of steps into a term, defined inductively as
\begin{align*}
\alpha ::= & \; {\cdot} \mid \lambda ; \alpha \mid @_1 ; \alpha \mid @_2 ; \alpha \mid \underline f_i ; \alpha \\
& {} \mid \tY ; \alpha \mid \textsf{if}_1 ; \alpha \mid \textsf{if}_2 ; \alpha \mid \textsf{if}_3 ; \alpha.
\end{align*}
The \emph{subterm of $M$ at a position $\alpha$}, denoted $M \mid \alpha$, is defined as
\begin{align*}
M \mid \cdot & = M \\
\lambda x. M \mid \lambda ; \alpha & = M \mid \alpha \\
M_1 M_2 \mid @_i ; \alpha & = M_i \mid \alpha \quad \text{for } i = 1,2 \\
\underline f(M_1,\dots,M_n) \mid \underline f_i ; \alpha & = M_i \mid \alpha \quad \text{for }i \leq n \\
\tY M \mid \tY ; \alpha & = M \mid \alpha \\
\tif{M_1 < 0}{M_2}{M_3} \mid \textsf{if}_i ; \alpha & = M_i \mid \alpha \quad \text{for } i = 1,2,3
\end{align*}
so that every subterm is located at a unique position, but not every position corresponds to a subterm (e.g. $x \, y \mid \lambda$ is undefined). 
A position such that $M\mid \alpha$ does exist is said to \emph{occur} in $M$. 
%\changed[lo]{
\emph{Substitution} of $N$ at position $\alpha$ in $M$, written $M[N/\alpha]$, is defined similarly.
For example, let 
\[
M = \lambda x \, y. y \, (\tif{x < 0}{y \, (\underline f (x))}{\underline 3})
\qquad
\alpha =\lambda ; \lambda ; @_2 ; \textsf{if}_2 ; @_2
\]
then 
\(
M[\tsample / \alpha] = \lambda x \, y. y \, (\tif{x < 0}{y \, \tsample}{\underline 3}).
\)
%}

Two subterms $N_1$ and $N_2$ of a term $M$, corresponding to positions $\alpha_1$ and $\alpha_2$, can overlap in a few different ways. 
If $\alpha_1$ is an initial segment (i.e.~prefix) of $\alpha_2$ (written as $\alpha_1 \leq \alpha_2$), then $N_2$ is also a subterm of $N_1$. If neither $\alpha_1 \leq \alpha_2$ nor $\alpha_1 \geq \alpha_2$, the positions are said to be \emph{disjoint}. 
The notion of disjointness is mostly relevant in that if $\alpha_1$ and $\alpha_2$ are disjoint, performing a substitution at $\alpha_1$ will leave the subterm at $\alpha_2$ unaffected.

@@

We are thus led to consider the reduction relation on skeletons (and positions in a skeleton), which can be extended from the definitions on terms in the obvious way, with $\tif{\skeletonPlaceholder < 0}{A}{B}$ reducing nondeterministically to both $A$ and $B$, and $\tsample$ reducing to $\skeletonPlaceholder$.
For example, we have 
\(
(\lambda x . \tif{x < 0}{x}{\skeletonPlaceholder}) \, \tsample
\to
(\lambda x . \tif{x < 0}{x}{\skeletonPlaceholder}) \, \skeletonPlaceholder
\to
\tif{\skeletonPlaceholder < 0}{\skeletonPlaceholder}{\skeletonPlaceholder}
\to
\skeletonPlaceholder
\).

Given a closed term $M$, let $L_0(M)$ be the set of pairs, the first element of which is a $\to$-reduction sequence of skeletons starting at $\mathit{Sk}(M)$, and the second of which is a position in the final skeleton of the reduction sequence. As with the traces from $I^{\mathbb N}$ used to pre-select samples to use in the standard sampling semantics, there will be (with one more caveat introduced later) a sample from $I$ pre-selected for each element of $L_0(M)$, which will then be used if a $\tsample$ reduction is ever performed at that position.

@@

\begin{example}[Labelling samples by terms]
Consider the terms
\begin{align*}
A[x] &= \tif{\tif{x>0}{I}{I} \, \tsample - 0.5 > 0}{0}{\Omega}\\
B &= \tif{\tsample - 0.5 > 0}{0}{\Omega}
\end{align*}
If terms rather than skeletons were used to label samples, the set of traces where $A[\tsample]$ terminates would be
\begin{align*}
\bigcup_{r \in [0,1]} \Big\{s \mid\ & s(A[\tsample], \mathsf{if}_1;\underline{-}_1;@_1;\mathsf{if}_1) = r, \\
& s(A[\tsample] \to A[\underline r] \to^* B, \mathsf{if}_1 ; \underline{-}_1) > 0.5 \Big\}.
\end{align*}

\lo{We have just defined $L_0(M)$, but what is $L_s(\cdot)$ (subscript $s$)? I assume it is the version of $L_0(M)$ that labels samples by terms.}

\akr{$L_s(M)$ is the set of potential positions (quotiented by $\sim^*$) where there is a $\tsample$. I don't want to move this example until after $L_s(M)$ has actually been defined, but also I don't want to imply that $I^{L_0(M)}$ is the actual entropy space too explicitly, so I'll just leave the set $s$ is restricted to implicit for now.}

\lo{The occurrence of $L_s(M)$ in the set comprehension below should be $L_0(M)$?}

\lo{I think we should write ``$[A[\tsample], A[\underline r], \ldots, B$'' in the line above as ``$A[\tsample] \to A[\underline r] \to^\ast B$''; similarly $\mathit{Sk}(A[\tsample]) \to \mathit{Sk}(A[\underline 0]) \to^\ast \mathit{Sk}(B)$ below.}
This is a rather unwieldy expression, but the crucial part is that $r$ occurs twice in the conditions on $s$: once as the value a sample must take, and once in the location of a sample. 
This set is unmeasurable, therefore the termination probability would not even be well-defined. 
Labelling samples by skeletons instead, this problem does not occur because there are only countably many skeleton, and at each step in a reduction sequence, only finitely many could have occurred yet. Although skeletal reduction sequences omit the information on what the results of sampling were, they still contain all the necessary information on how many, and which, reductions took place.

For this particular term, $\mathit{Sk}(A[\underline r])$ does not depend on the value of $r$, therefore the set where it terminates becomes simply
\begin{align*}
\big\{s \mid s(\mathit{Sk}(A[\tsample]) \to \mathit{Sk}(A[\underline 0]) \to^* \mathit{Sk}(B), \mathsf{if}_1 ; \underline{-}_1) > 0.5 \big\}
\end{align*}
which is measurable.
\qed
\end{example}

@@

The reflexive transitive closure $\sim^*$ of this relation is used to define the set of \emph{potential positions} $L(M) = L_0(M) / \sim^*$, and each equivalence class can be considered as the same position as it may occur across multiple reachable skeletons. 
If $(N,\alpha) \sim^* (O,\beta)$, then $N \mid \alpha$ and $O \mid \beta$ both have the same shape (i.e.~they're either both the placeholder $\skeletonPlaceholder$, both variables, both applications, both $\tsample$s etc.), therefore it's well-defined to talk of the set of potential positions where there is a $\tsample$, $L_s(M)$. 
Formally $L_s(M) := \{[(X, \alpha)]_{\sim^\ast_M} : X \mid \alpha = \tsample\}$.
The new sample space is then defined as $I^{L_s(M)}$, with the Borel $\sigma$-algebra and product measure.
\changed[lo]{Since $I^{L_s(M)}$ is a countable product, the measure space is well-defined \citep[Cor.~2.7.3]{AshDD00}.}

@@

Unlike in the purely call-by-value case, the version of the reduction relation that takes into account samples is still a general relation rather than a function, so it is denoted ``$\Rightarrow$'' instead of ``$\red$'', and it relates $\biguplus_{M \in \Lambda_0} I^{L_s(M)}$ to itself.
\changed[lo]{We write an element of $\biguplus_{M \in \Lambda_0} I^{L_s(M)}$ as $(M', s)$ where the \emph{term} $M' \in \Lambda^0$ and $s \in  I^{L_s(M')}$.}
\begin{align*}
& (M,s) \Rightarrow (N,s \circ i(M \to N)) \text{ if $M \to N$ at $\alpha$ and either} \\
& \qquad \text{the redex is not $\tsample$, or} \\
& \qquad \text{$M \mid \alpha = \tsample$ and $N = M[\underline{s(\mathit{Sk}(M),\alpha)}/\alpha]$}
\end{align*}

>>>>

A \emph{position} is a finite sequence of steps into a term, defined inductively as
\begin{align*}
\alpha ::= & \; {\cdot} \mid \lambda ; \alpha \mid @_1 ; \alpha \mid @_2 ; \alpha \mid \underline f_i ; \alpha \\
& {} \mid \tY ; \alpha \mid \textsf{if}_1 ; \alpha \mid \textsf{if}_2 ; \alpha \mid \textsf{if}_3 ; \alpha.
\end{align*}
The \emph{subterm of $M$ at a position $\alpha$}, denoted $M \mid \alpha$, is defined as
\begin{align*}
M \mid \cdot & = M \\
\lambda x. M \mid \lambda ; \alpha & = M \mid \alpha \\
M_1 M_2 \mid @_i ; \alpha & = M_i \mid \alpha \quad \text{for } i = 1,2 \\
\underline f(M_1,\dots,M_n) \mid \underline f_i ; \alpha & = M_i \mid \alpha \quad \text{for }i \leq n \\
\tY M \mid \tY ; \alpha & = M \mid \alpha \\
\tif{M_1 < 0}{M_2}{M_3} \mid \textsf{if}_i ; \alpha & = M_i \mid \alpha \quad \text{for } i = 1,2,3
\end{align*}
so that every subterm is located at a unique position, but not every position corresponds to a subterm (e.g. $x \, y \mid \lambda$ is undefined). 
A position such that $M\mid \alpha$ does exist is said to \emph{occur} in $M$. 
%\changed[lo]{
\emph{Substitution} of $N$ at position $\alpha$ in $M$, written $M[N/\alpha]$, is defined similarly.
For example, let 
\[
M = \lambda x \, y. y \, (\tif{x < 0}{y \, (\underline f (x))}{\underline 3})
\quad
\alpha =\lambda ; \lambda ; @_2 ; \textsf{if}_2 ; @_2
\]
then 
\(
M[\tsample / \alpha] = \lambda x \, y. y \, (\tif{x < 0}{y \, \tsample}{\underline 3}).
\)
%}

Two subterms $N_1$ and $N_2$ of a term $M$, corresponding to positions $\alpha_1$ and $\alpha_2$, can overlap in a few different ways. 
If $\alpha_1$ is a prefix of $\alpha_2$ (written as $\alpha_1 \leq \alpha_2$), then $N_2$ is also a subterm of $N_1$. If neither $\alpha_1 \leq \alpha_2$ nor $\alpha_1 \geq \alpha_2$, the positions are said to be \emph{disjoint}. 
The notion of disjointness is mostly relevant in that if $\alpha_1$ and $\alpha_2$ are disjoint, performing a substitution at $\alpha_1$ will leave the subterm at $\alpha_2$ unaffected.

@@

We are thus led to consider the reduction relation on skeletons (and positions in a skeleton), which can be extended from the definitions on terms in the obvious way, with $\tif{\skeletonPlaceholder < 0}{A}{B}$ reducing nondeterministically to both $A$ and $B$, and $\tsample$ reducing to $\skeletonPlaceholder$, \changed[lo]{and $(\lambda x.A) \, \skeletonPlaceholder$ reducing to $A[\skeletonPlaceholder / x]$}.
For example, we have 
\(
(\lambda x . \tif{x < 0}{x}{\skeletonPlaceholder}) \, \tsample
\to
(\lambda x . \tif{x < 0}{x}{\skeletonPlaceholder}) \, \skeletonPlaceholder
\to
\tif{\skeletonPlaceholder < 0}{\skeletonPlaceholder}{\skeletonPlaceholder}
\to
\skeletonPlaceholder
\).

Given a closed term $M$, let $L_0(M)$ be the set of pairs, the first element of which is a $\to$-reduction sequence of skeletons starting at $\mathit{Sk}(M)$, and the second of which is a position in the final skeleton of the reduction sequence. 
As with the traces from $I^{\mathbb N}$ used to pre-select samples to use in the standard sampling semantics, 
%there will be (with one more caveat introduced later) a sample from $I$ pre-selected for each element of $L_0(M)$, which will then be used if a $\tsample$ reduction is ever performed at that position.
\changed[lo]{modified traces, which are elements of $I^{L_0(M)}$ (with one more caveat introduced later)}, will be used to pre-select a sample from $I$ for each element of $L_0(M)$, which will then be used if a $\tsample$ reduction is ever performed at that position.

@@

\begin{example}[Labelling samples by terms]
Consider the terms
\begin{align*}
A[x] &= \tif{\tif{x>0}{I}{I} \, \tsample - 0.5 > 0}{0}{\Omega}\\
B &= \tif{\tsample - 0.5 > 0}{0}{\Omega}
\end{align*}
If terms rather than skeletons were used to label samples, the set of \changed[akr]{``traces''} \lo{Readers may not realise that traces here means elements of $I^{L_0(M)}$ (but with samples labelled by terms). I propose replacing the highlighted ``traces'' by ``modified traces'' (which is introduced two paragraphs before). I think this might be a little clearer, but feel free to change back.} where $A[\tsample]$ terminates would be
\begin{align*}
\bigcup_{r \in [0,1]} \Big\{s \mid\ & s(A[\tsample], \mathsf{if}_1;\underline{-}_1;@_1;\mathsf{if}_1) = r, \\
& s(A[\tsample] \to A[\underline r] \to^* B, \mathsf{if}_1 ; \underline{-}_1) > 0.5 \Big\}.
\end{align*}

\lo{We have just defined $L_0(M)$, but what is $L_s(\cdot)$ (subscript $s$)? I assume it is the version of $L_0(M)$ that labels samples by terms.} 
\akr{$L_s(M)$ is the set of potential positions (quotiented by $\sim^*$) where there is a $\tsample$. I don't want to move this example until after $L_s(M)$ has actually been defined, but also I don't want to imply that $I^{L_0(M)}$ is the actual entropy space too explicitly, so I'll just leave the set $s$ is restricted to implicit for now.}
\lo{Noted.}

\iffalse
\lo{The occurrence of $L_s(M)$ in the set comprehension below should be $L_0(M)$?}

\lo{I think we should write ``$[A[\tsample], A[\underline r], \ldots, B$'' in the line above as ``$A[\tsample] \to A[\underline r] \to^\ast B$''; similarly $\mathit{Sk}(A[\tsample]) \to \mathit{Sk}(A[\underline 0]) \to^\ast \mathit{Sk}(B)$ below.}
\fi
This is a rather unwieldy expression, but the crucial part is that $r$ occurs twice in the conditions on $s$: once as the value a sample must take, and once in the location of a sample. 
This set is unmeasurable, therefore the termination probability would not even be well-defined. 
Labelling samples by skeletons instead, this problem does not occur because there are only countably many skeleton, and at each step in a reduction sequence, only finitely many could have occurred yet. Although skeletal reduction sequences omit the information on what the results of sampling were, they still contain all the necessary information on how many, and which, reductions took place.

For this particular term, $\mathit{Sk}(A[\underline r])$ does not depend on the value of $r$, therefore the set where it terminates becomes simply the following, which is measurable.
\begin{align*}
\big\{s \mid s(\mathit{Sk}(A[\tsample]) \to \mathit{Sk}(A[\underline 0]) \to^* \mathit{Sk}(B), \mathsf{if}_1 ; \underline{-}_1) > 0.5 \big\}
\end{align*}
\end{example}

@@

The reflexive transitive closure $\sim^*$ of this relation is used to define the set of \emph{potential positions} $L(M) = L_0(M) / \sim^*$, and each equivalence class can be considered as the same position as it may occur across multiple reachable skeletons. 
If $(N,\alpha) \sim^* (O,\beta)$, then $N \mid \alpha$ and $O \mid \beta$ both have the same shape (i.e.~they're either both the placeholder $\skeletonPlaceholder$, both variables, both applications, both $\tsample$s etc.), therefore it's well-defined to talk of the set of potential positions where there is a $\tsample$, $L_s(M)$. 
Formally $L_s(M) := \{[(X, \alpha)]_{\sim^\ast_M} : X \mid \alpha = \tsample\}$.
The new sample space is then defined as $I^{L_s(M)}$, with the Borel $\sigma$-algebra and product measure.
Since $I^{L_s(M)}$ is a countable product, the measure space is well-defined \citep[Cor.~2.7.3]{AshDD00}.

@@

\begin{definition}
Unlike in the purely call-by-value case, the version of the reduction relation that takes into account samples is still a general relation rather than a function, so it is denoted ``$\Rightarrow$'' instead of ``$\red$'', and it relates $\biguplus_{M \in \Lambda_0} I^{L_s(M)}$ to itself.
\changed[lo]{We write an element of $\biguplus_{M \in \Lambda_0} I^{L_s(M)}$ as $(M', s)$ where the \emph{term} $M' \in \Lambda^0$ and $s \in  I^{L_s(M')}$.}
\begin{align*}
& (M,s) \Rightarrow (N,s \circ i(M \to N)) \text{ if $M \to N$ at $\alpha$ and either} \\
& \qquad \text{the redex is not $\tsample$, or} \\
& \qquad \text{$M \mid \alpha = \tsample$ and $N = M[\underline{s(\mathit{Sk}(M),\alpha)}/\alpha]$}
\end{align*}
\end{definition}

### 2021-01-05 LICS21/ranking.tex
Sec. VI. Ex. VI.2

\begin{example}
\label{ex:tY finite does not imply t finite}
The term $M = \Xi \, (\lambda x. x+1)$ where
\[
\Xi := \tY \lambda f n. \tif{\tsample - 0.5 < 0}{n \, \underline 0}{f (\lambda x. n (n x))}
\] 
terminates with only 2 $\tY$-reductions on average, i.e.,~$\expect{T_M^\tY} = 2$, but applies the increment function $2^n$ times with probability $2^{-n-1}$ for $n \geq 0$, which diverges, i.e.,~$\expect{T_M} = \infty$.
\end{example}

@@

The definition of rankability is also inconvenient for syntactic sugar. It could be useful, for example, to define $M \oplus_p N = \tif{\tsample - p < 0} M N$, where $M \oplus_p N$ reduces to $M$ or $N$, depending on the first value of $s \in \entrosp$, with probability $p$ resp.~$(1-p)$. Technically though, it reduces first to $\tif{\underline r - p < 0} M N$ for all $r \in I$, so those terms all need values of the ranking function too.

%\paragraph{}
In both of these cases, there are only some values of the ranking function that are semantically important. 
\begin{definition}
Define a \emph{sparse ranking function} on a closed term $M$ to be a partial function $f : Rch(M) \rightharpoonup \mathbb R$ such that
\begin{itemize}
    \item $f(N) \geq 0$ for all $N$ where $f$ is defined.
    \item $f$ is defined at $M$.
    \item For any $N$ in the domain of definition of $f$, evaluation of $N$ will eventually \akr{It would be nicer for this to be only almost certain, to make this more neatly in-between rankability and ast, but that doesn't actually work because of terms reachable with 0 probability. Of course, astness doesn't imply rankability, so any definition of sparse ranking function that includes all Y-past terms wouldn't always extend to a total ranking function.} reach some $O$ which is either a value or in the domain of definition of $f$, and $f(N) \geq \mathbb E[f(O) + \text{ the number of $\tY$-reduction steps from $N$ to $O$}]$ (where $f(O)$ is taken to be 0 if $O$ is a value outside of the domain of $f$).
\end{itemize}
\end{definition}
A sparse ranking function that is total is just a ranking function. Providing a sparse ranking function is essentially part way between providing a ranking function and directly proving almost sure termination.

@@

%\paragraph{}
With this abbreviation, the geometric distribution example from earlier can be written as $(\tY \lambda f n.\ n \oplus_{0.5} f(n+1)) \, \underline 0$. It is then easy to see that the following is a sparse ranking function:
\begin{align*}
(\tY \lambda f n.\ n \oplus_{0.5} f(n+1))\ N & \mapsto 2 \\
\underline i \oplus_{0.5} (\tY \lambda f n.\ n \oplus_{0.5} f(n+1))\ (\underline i + 1) & \mapsto 1 \\
\underline i & \mapsto 0.
\end{align*}

>>>>

\begin{example}
\label{ex:tY finite does not imply t finite}
The term $M = \Xi \, (\lambda x. x+1)$ where
\[
\Xi := \tY \lambda f \, n. \tif{\tsample - 0.5 < 0}{n \, \underline 0}{f (\lambda x. n (n \, x))}
\] 
terminates with only 2 $\tY$-reductions on average, i.e.,~$\expect{T_M^\tY} = 2$, but applies the increment function $2^n$ times with probability $2^{-n-1}$ for $n \geq 0$, which diverges, i.e.,~$\expect{T_M} = \infty$.
\end{example}

@@

The definition of rankability is also inconvenient for syntactic sugar. It could be useful, for example, to define $M \oplus_p N := \tif{\tsample - p < 0} M N$, where $M \oplus_p N$ reduces to $M$ or $N$, depending on the first value of $s \in \entrosp$, with probability $p$ resp.~$(1-p)$. Technically though, it reduces first to $\tif{\underline r - p < 0} M N$ for all $r \in I$, so those terms all need values of the ranking function too.

%\paragraph{}
In both of these cases, there are only some values of the ranking function that are semantically important. 
\begin{definition}
Define a \emph{sparse ranking function} on a closed term $M$ to be a partial function $f : Rch(M) \rightharpoonup \mathbb R$ such that
\begin{itemize}
    \item $f(N) \geq 0$ for all $N$ where $f$ is defined.
    \item $f$ is defined at $M$.
    \item For any $N$ in the domain of definition of $f$, evaluation of $N$ will eventually \akr{It would be nicer for this to be only almost certain, to make this more neatly in-between rankability and AST, but that doesn't actually work because of terms reachable with 0 probability. Of course, ASTness doesn't imply rankability, so any definition of sparse ranking function that includes all Y-PAST terms wouldn't always extend to a total ranking function.} reach some $O$ which is either a value or in the domain of definition of $f$, and $f(N) \geq \mathbb E[f(O) + \text{ the number of $\tY$-reduction steps from $N$ to $O$}]$ (where $f(O)$ is taken to be 0 if $O$ is a value outside of the domain of $f$).
\end{itemize}
\end{definition}
A sparse ranking function that is total is just a ranking function. Providing a sparse ranking function is essentially part way between providing a ranking function and directly proving almost sure termination.

@@

%\paragraph{}
With this abbreviation, the geometric distribution example from earlier can be written as $(\tY \lambda f n.\ n \oplus_{0.5} f(n+1)) \, \underline 0$. It is then easy to see that the following is a sparse ranking function:
\begin{align*}
(\tY \lambda f n.\ n \oplus_{0.5} f(n+1))\, N & \mapsto 2 \\
\underline i \oplus_{0.5} (\tY \lambda f n.\ n \oplus_{0.5} f(n+1))\, (\underline i + 1) & \mapsto 1 \\
\underline i & \mapsto 0.
\end{align*}
